---
title: "Woody Biomass Stocks and Fluxes in the BCNM from Plot Data"
author:
  - Camille Piponiot:
      email: camille.piponiot-laroche@cirad.fr
      institute: [fs]
  - Helene C. Muller-Landau:
      institute: [stri] 

institute:
  - fs: Cirad, Université de Montpellier, UR Forests and Societies, 34980 Montferrier-sur-Lez, France
  - stri: Forest Global Earth Observatory, Smithsonian Tropical Research Institute, Apartado Postal 0843-03092, Panamá, Republic of Panamá

output: 
  bookdown::word_document2:
    pandoc_args:
      - '--lua-filter=common/scholarly-metadata.lua'
      - '--lua-filter=common/author-info-blocks.lua'
    reference_docx: common/word-styles-ref-01.docx
  fig_caption: yes
bibliography: references.bib
csl: common/new-phytologist.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, results='hide')
# list of package dependencies
req_packages <-
  c("rdryad",
    "data.table",
    "ggplot2",
    "utils",
    "truncnorm",
    "ggpubr",
    "readxl")

# packages that are not yet installed on the computer
ins_packages <-  req_packages[!(req_packages %in% rownames(installed.packages()))]

# install missing packages
if (length(ins_packages) > 0) 
  install.packages(ins_packages)

# load packages
library(ggplot2)

# source internal functions 
source("functions.R")
```

```{r run-main-script}
if (!file.exists("data/data-main-script.rda"))
  source("main-script.R") else
    load("data/data-main-script.rda")
```

# Introduction

Biomass is the mass of all living things in an ecosystem; this amount is usually measured on oven-dried material, thus excluding water. The biomass of a forest is dominated by the aerial parts of the trees, mainly the trunks and branches [ref]. It was rarely considered in ecology until about 30 years ago, but has recently received much attention due to increased interest in the global carbon cycle and the crucial role of forests in this cycle [ref]. This has led to an increase in the number of papers calculating forest biomass stocks and biomass fluxes associated with tree growth and mortality.

The traditional method of estimating tree biomass is to measure the basic characteristics (such as diameter) of all trees in a given area, and derive an estimate of biomass. On Barro Colorado Island, a 50-ha plot was established in 1980 to regularly measure all stems from 1 cm in diameter upward, allowing for an extensive estimate of woody biomass in the plot and its changes over time. In this chapter, we will explore the results of 35 years of measurements on the plot, focusing on aboveground biomass stocks and fluxes: how to estimate them, their temporal and spatial variation on the plot, and their distribution into size and functional groups. We will then briefly present other biomass stocks and fluxes, including below-ground biomass of trees, other plants (lianas, others?)

```{r fig1-allom, fig.height = 4, fig.width = 10,fig.cap="AGB estimates using different allometries (shown by different colors). (a) AGB estimated per individual tree as a function of DBH for trees with different wood densities (plain lines: wood density = 0.4; dotted lines: wood density = 0.8); (b) plot-level AGB for all censuses 1985-2015. The allometries used were: Chave et al.'s (2005) allometry for wet forest stands, without tree height (red lines) or using the generic height allometry of Martinez Caño et al. (2019) developed at BCI (blue lines); or Chave et al.'s (2014) allometry, without tree height (equation 7 - green lines) or with tree height (equation 4 - purple lines) using the generic height allometry of Martinez Caño et al. (2019)."}
#### Figure 1 - AGB estimation and potential sources of uncertainty ####

# create data table for illustrating different allometric equations at the
# individual tree level
dfallom <- data.table::data.table(expand.grid(dbh = 1:200, wd = c(0.4, 0.8)))

# Chave et al 2014 allometric equation, no height information
dfallom[, chave14 := agb_bci(dbh = dbh, wd = wd, method = "chave14")]

# Chave et al 2014 allometric equation, tree height from Martinez Cano et al., 2019
dfallom[, chave14_h := agb_bci(dbh = dbh, wd = wd, method = "chave14", use_height_allom = TRUE)]

# Chave et al 2005 allometric equation, no height information
dfallom[, chave05 := agb_bci(dbh = dbh, wd = wd, method = "chave05")]

# Chave et al 2005 allometric equation, tree height from Martinez Cano et al., 2019
dfallom[, chave05_h := agb_bci(dbh = dbh, wd = wd, method = "chave05", use_height_allom = TRUE)]

dfallom <- data.table::melt(dfallom, measure.vars = grep("chave", colnames(dfallom)), 
                            variable.name = "method", value.name = "agb")

# change order of method levels to have the legend in the same order as the lines in the figure
dfallom$method <- factor(dfallom$method, levels = c("chave05", "chave14", "chave05_h", "chave14_h"))
levels(dfallom$method) <- list("Chave 2005" = "chave05",
                               "Chave 2014" = "chave14",
                               "Chave 2005 + height allom" = "chave05_h",
                               "Chave 2014 + height allom" = "chave14_h")

fig1a_allom_ind <-
  ggplot(dfallom, aes(
    x = dbh,
    y = agb,
    colour = method,
    linetype = as.factor(wd)
  )) +
  geom_line() +
  theme_classic() + 
  expand_limits(x = 0, y = 0) +   
  scale_x_continuous(expand = c(0, 0)) + 
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = "Stem diameter (cm)",
       y = "Estimated aboveground biomass of trees (Mg)",
       colour = "Allometry used",
       lty = "Wood density") 
fig1_leg <- ggpubr::as_ggplot(ggpubr::get_legend(fig1a_allom_ind))

# subset data frame to get only biomass estimates without corrections
df_agb <- subset(df_plot, year >= 1985 & variable == "agb" & !grepl("tion", method))

# change order of method levels to have the legend in the same order as the lines in the figure
df_agb$method <- factor(df_agb$method,
                        levels = paste0(rep(c("Chave 2005", "Chave 2014"), 2),
                                        rep(c("", " + height allom"), each = 2)))

fig1b_allom_plot <- ggplot(df_agb) +
  geom_line(aes(x = year, y = value, color = method)) +
  theme_classic() +
  expand_limits(y = 0) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = "Census year",
       y = "Estimated plot aboveground biomass (Mg/ha)",
       colour = "Allometry used") +
  theme(legend.position = "none")

ggpubr::ggarrange(fig1a_allom_ind + theme(legend.position = "none"), 
                  fig1b_allom_plot, fig1_leg, 
                  labels = c("a", "b", ""), ncol = 3, widths = c(2,2,1))
```

```{r fig2-correc, fig.height = 4, fig.width = 12}
# Figure 2 - Corrections ####
# fig2a - illustration with one problematic tree
df_929 <- data.table::melt(df_929, measure.vars = c("dbh", "hom"), id.vars = "year")
levels(df_929$variable) <- c("Measured diameter (cm)", "Height of measurement (m)")
fig2a_corr_tree <- ggplot(df_929, aes(x = year, y = value, color = variable)) +
  geom_point() + 
  labs(y = "", x = "Census year", title = "Tree #929") +
  facet_wrap(~variable, ncol = 1, scales = "free_y", 
             strip.position = "left")+
  theme_classic() +
  theme(legend.position = "none", 
        strip.placement = "outside", 
        strip.background = element_blank()) 

# fig2b - corrections on plot AGB
dfcorr <- subset(df_plot, year >= 1985 &
                   (grepl("corr", method) | method == "Chave 2014") & 
                   variable == "agb")
levels(dfcorr$method)[levels(dfcorr$method) == "Chave 2014"] <- "No correction"

fig2b_corr_agb <-
  ggplot(dfcorr) +
  geom_line(aes(x = year, y = value, color = method)) +
  theme_classic() +
  labs(x = "Census year",
       y = "Estimated plot aboveground biomass (Mg/ha)",
       colour = "Correction applied") + 
  theme(legend.position = c(.75, .55))

fig2c_corr_awp <-
  ggplot(subset(df_plot, year >= 1985 & year < 2015 & variable == "awp")) +
  geom_line(aes(x = year, y = value, color = method)) +
  theme_classic() +
  labs(x = "Census year",
       y = "Estimated plot aboveground woody \nproductivity (Mg/ha/yr)",
       colour = "Correction applied") + 
  theme(legend.position = c(.5, .75))

ggpubr::ggarrange(fig2a_corr_tree, fig2b_corr_agb, fig2c_corr_awp, 
                  nrow = 1, labels = "auto")
```

```{r fig3-groups, fig.height = 4, fig.width = 8, fig.cap="Stacked barplots showing the proportion of AGB and AWP (%) in different groups. (a) Proportion of AGB (left) and AWP (right) by size class (from blue to red).  Proportion of AGB (left) and AWP (right) by plant functional type as defined in Ruger et al. (2020.)"}
# Figure 3 - AGB per group ####

# # estimate percentage of each size class or PFT group per variable
df_size[, `:=`(agb = agb/sum(agb), awp = awp/sum(awp))]
# revert order of size classes to have bigger trees above
df_size$size <- factor(df_size$size, levels = rev(levels(df_size$size)))
# 
df_pft[, `:=`(agb = agb/sum(agb), awp = awp/sum(awp))]

df_size_bis <- data.table::melt(df_size, measure.vars = c("agb", "awp"))
# change variable labels (AGB, AWP) to upper case
levels(df_size_bis$variable) <- toupper(levels(df_size_bis$variable))
# figure 3a bis
fig3a_bis_groups_size <- ggplot(df_size_bis) + 
  geom_col(aes(x = variable, y = value*100, fill = size)) + 
  labs(x = "", y = "Proportion (%)", fill = "Diameter\nclass (cm)") +
  scale_y_continuous(expand = c(0, 0)) +  
  scale_fill_brewer(palette = "RdYlBu") +
  theme_classic()

df_pft_bis <- data.table::melt(df_pft, measure.vars = c("agb", "awp"))
# change variable labels (AGB, AWP) to upper case
levels(df_pft_bis$variable) <- toupper(levels(df_pft_bis$variable))
# figure 3a bis
fig3b_bis_groups_pft <- ggplot(df_pft_bis) + 
  geom_col(aes(x = variable, y = value*100, fill = PFT)) + 
  labs(x = "", y = "Proportion (%)", fill = "PFT") +
  scale_y_continuous(expand = c(0, 0)) +  
  scale_fill_manual(values = c("gold", "darkorange", "seagreen", "royalblue", "orchid")) +
  theme_classic()

ggpubr::ggarrange(fig3a_bis_groups_size, fig3b_bis_groups_pft, ncol = 2, labels = "auto")
```

# Plot-based tree aboveground biomass methods

* How to measure/calculate aboveground biomass stocks 

> The aboveground biomass can be measured directly by harvesting, drying, and weighing the trees. Such destructive methods are expensive, and have never been done on BCI due to restrictions on destructive sampling. Aboveground biomass on BCI was therefore inferred by applying allometric equations based on harvests elsewhere. These allometric equations estimate the biomass of individual trees from their diameters, and sometimes their heights and species wood densities. The estimated aboveground biomass is then summed across all trees to obtain the total aboveground biomass of the stand. 
    
* Estimating individual tree biomass from allometric equations

    - Estimating of tree biomass from allometric equations requires choosing an equation 
    
    - Early equations just based on diameter and few trees (e.g., as used in Chave et al. 2003 paper)
    
    - Later equations incorporate tree height, or environmental factors that are proxies for height, as well as wood density (cite papers giving these equations, and papers using them)
    
    - For sites where heights of all trees not available (like BCI), this can then also mean choosing a height allometry equation. (cite Martinez-Cano height allometry equation, data)
    
    - Wood density assignments also an issue – most often from large taxonomic databases, but there is considerable variation within species, both among and within sites, so more local values better (cite wood density papers, data). 
    
    - When converting biomass to carbon, additional issue of carbon fraction, which is variable and little-quantified 
    
    - Importance of allometry for biomass stocks at plot level (figure 1, part)
    
* Calculating biomass fluxes from allometric equations 

    - Mortality flux – simple, and as corrected by Kohyama
    
    - Productivity flux – recruitment and growth – simple, and as corrected 
    
    - Note that this woody productivity underestimates true wood production because it misses branch turnover, which is also missing from mortality flux.  
    
    - Net flux / net change – simple in principle….
    
* Complications of nonstandard height of measurement, data QAQC

    - Estimates of biomass stocks and fluxes highly sensitive to procedures applied to deal with nonstandard heights of measurement and outlier values for change in diameter
    
    - Muller-Landau book chapter
    
    - Cushman taper correction 
    
    - Also specific issues with early BCI plot data – measurements of buttressed trees, small stems, no tags on stems so multi-stemmed dynamics problematic
    
    - These are why no two papers give the same values for BCI biomass stocks and fluxes!
    
    - Importance of QAQC methods for biomass productivity at plot level (figure 1, part)
