---
title: "Woody Biomass Stocks and Fluxes in the BCNM from Plot Data"
author:
  - Camille Piponiot:
      email: camille.piponiot-laroche@cirad.fr
      institute: [fs]
  - Helene C. Muller-Landau:
      institute: [stri] 

institute:
  - fs: Cirad, Université de Montpellier, UR Forests and Societies, 34980 Montferrier-sur-Lez, France
  - stri: Forest Global Earth Observatory, Smithsonian Tropical Research Institute, Apartado Postal 0843-03092, Panamá, Republic of Panamá

output: 
  bookdown::word_document2:
    pandoc_args:
      - '--lua-filter=common/scholarly-metadata.lua'
      - '--lua-filter=common/author-info-blocks.lua'
    reference_docx: common/word-styles-ref-01.docx
  fig_caption: yes
bibliography: references.bib
csl: common/new-phytologist.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, results = "hide")
# list of package dependencies
req_packages <-
  c("rdryad",
    "data.table",
    "ggplot2",
    "utils",
    "truncnorm",
    "ggpubr",
    "readxl", 
    "raster", 
    "sf")

# packages that are not yet installed on the computer
ins_packages <-  req_packages[!(req_packages %in% rownames(installed.packages()))]

# install missing packages
if (length(ins_packages) > 0) 
  install.packages(ins_packages)

# load packages
library(ggplot2)

# source internal functions 
source("functions.R")
```

```{r run-main-script}
if (!file.exists("data/data-main-script.rda"))
  source("main-script.R") else
    load("data/data-main-script.rda")
```

# Introduction

Biomass is the mass of all living things in an ecosystem, and is usually measured on oven-dried material, thus excluding water. The biomass of a forest is dominated by the aerial parts of the trees, mainly the trunks and branches [ref]. It was rarely considered in ecology until about 30 years ago, but has recently received much attention due to increased interest in the global carbon cycle and the crucial role of forests in this cycle [ref]. This has led to an increase in the number of papers calculating forest biomass stocks and biomass fluxes associated with tree growth and mortality [ref].

The traditional method of estimating tree biomass is to measure the basic characteristics (such as diameter) of all trees in a given area, and derive an estimate of biomass. On Barro Colorado Island, a 50-ha plot was established in 1980 to regularly measure all stems from 1 cm in diameter upward, allowing for an extensive estimate of woody biomass in the plot and its changes over time. In this chapter, we will explore the results of 35 years of measurements on the plot, focusing on aboveground biomass stocks and fluxes: how to estimate them, their temporal and spatial variation on the plot, and their distribution into size and functional groups. We will then briefly present other biomass stocks and fluxes, including below-ground biomass of trees, other plants (lianas, others?)

# Plot-based tree aboveground biomass methods

> add something about plot layout, what trees are measured (eg dbh threshold?)

* How to measure/calculate aboveground biomass stocks 

The aboveground biomass can be measured directly by harvesting, drying, and weighing the trees. Such destructive methods are expensive, and have never been done on BCI due to restrictions on destructive sampling. Aboveground biomass on BCI was therefore inferred by applying allometric equations based on harvests elsewhere. These allometric equations estimate the biomass of individual trees from their diameters, and sometimes their heights and species wood densities. The estimated aboveground biomass is then summed across all trees to obtain the total aboveground biomass of the stand.

* Estimating individual tree biomass from allometric equations

Estimating the biomass of an individual tree from allometric equations is not a trivial task, as it requires many choices and approximations. Early allometric equations were calibrated with only a few trees, usually at few sites, and used only diameter as a predictor variable [e.g., as used in @Chave2003]. Later developed equations include a larger number of sites spread across the tropics, incorporate tree height, or environmental factors that are proxies for height, and wood density [@Chave2005;@Chave2014]. Tree height is more difficult to measure than diameter, especially in dense vegetation: for sites where heights of all trees are not available (such as BCI), then a height allometry equation can be chosen [e.g., @MartinezCano2019 which was calibrated to BCI]. Wood density is most often assigned by species from large taxonomic databases [e.g., @Zanne2009], but there is considerable variation within species, both among and within sites [@Muller-Landau2004]. Finally, the conversion of biomass to carbon is also subject to error, as the fraction of carbon in biomass is variable and poorly quantified [@Thomas2012]. The set of choices when estimating tree biomass can therefore have a significant impact on the estimate of plot biomass stocks (Figure \@ref(fig:fig1-allom)a). 

```{r fig1-allom, fig.height = 4, fig.width = 10,fig.cap="AGB estimates using different allometries (shown by different colors). (a) aboveground biomass (AGB) estimated per individual tree as a function of DBH for trees with different wood densities (plain lines: wood density = 0.4; dotted lines: wood density = 0.8); (b) plot-level AGB for all censuses 1985-2015. The allometries used were: the allometry from @Chave2005 for moist forest stands, without tree height (red lines) or using the generic height allometry of @MartinezCano2019 developed at BCI (blue lines); or the allometry from @Chave2014, without tree height (equation 7 - green lines) or with tree height (equation 4 - purple lines) using the generic height allometry of @MartinezCano2019."}
#### Figure 1 - AGB estimation and potential sources of uncertainty ####

# create data table for illustrating different allometric equations at the
# individual tree level
dfallom <- data.table::data.table(expand.grid(dbh = 1:200, wd = c(0.4, 0.8)))

# Chave et al 2014 allometric equation, no height information
dfallom[, chave14 := agb_bci(dbh = dbh, wd = wd, method = "chave14")]

# Chave et al 2014 allometric equation, tree height from Martinez Cano et al., 2019
dfallom[, chave14_h := agb_bci(dbh = dbh, wd = wd, method = "chave14", use_height_allom = TRUE)]

# Chave et al 2005 allometric equation, no height information
dfallom[, chave05 := agb_bci(dbh = dbh, wd = wd, method = "chave05")]

# Chave et al 2005 allometric equation, tree height from Martinez Cano et al., 2019
dfallom[, chave05_h := agb_bci(dbh = dbh, wd = wd, method = "chave05", use_height_allom = TRUE)]

dfallom <- data.table::melt(dfallom, measure.vars = grep("chave", colnames(dfallom)), 
                            variable.name = "method", value.name = "agb")

# change order of method levels to have the legend in the same order as the lines in the figure
dfallom$method <- factor(dfallom$method, levels = c("chave05", "chave14", "chave05_h", "chave14_h"))
levels(dfallom$method) <- list("Chave 2005" = "chave05",
                               "Chave 2014" = "chave14",
                               "Chave 2005 + height allom" = "chave05_h",
                               "Chave 2014 + height allom" = "chave14_h")

fig1a_allom_ind <-
  ggplot(dfallom, aes(
    x = dbh,
    y = agb,
    colour = method,
    linetype = as.factor(wd)
  )) +
  geom_line() +
  theme_classic() + 
  expand_limits(x = 0, y = 0) +   
  scale_x_continuous(expand = c(0, 0)) + 
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = "Stem diameter (cm)",
       y = "Estimated aboveground biomass of trees (Mg)",
       colour = "Allometry used",
       lty = "Wood density") 
fig1_leg <- ggpubr::as_ggplot(ggpubr::get_legend(fig1a_allom_ind))

# subset data frame to get only biomass estimates without corrections
df_agb <- subset(df_plot, variable == "agb" & !grepl("\\+", method))

# change order of method levels to have the legend in the same order as the lines in the figure
df_agb$method <- factor(df_agb$method, levels = c("chave05", "chave14", "chave05_h", "chave14_h"))
levels(df_agb$method) <- list("Chave 2005" = "chave05",
                              "Chave 2014" = "chave14",
                              "Chave 2005 + height allom" = "chave05_h",
                              "Chave 2014 + height allom" = "chave14_h")

fig1b_allom_plot <- ggplot(df_agb, aes(x = year, y = value, color = method)) +
  geom_point() + geom_line(lty = 2) + 
  theme_classic() +
  expand_limits(y = 0) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = "Census year",
       y = "Estimated plot aboveground biomass (Mg/ha)",
       colour = "Allometry used") +
  theme(legend.position = "none")

ggpubr::ggarrange(fig1a_allom_ind + theme(legend.position = "none"), 
                  fig1b_allom_plot, fig1_leg, 
                  labels = c("a", "b", ""), ncol = 3, widths = c(2,2,1))
```

* Calculating biomass fluxes from allometric equations 

In addition to biomass stocks, ecologists are also interested in biomass dynamics: changes in biomass over time, which is the net difference between biomass increase due to tree productivity and biomass loss due to tree mortality. Annual mortality is calculated as the biomass of all dead trees between two censuses, divided by the time period between the two censuses. Annual woody productivity is the annualized sum of the increase in biomass of living trees between two censuses, and the biomass of recruited trees, i.e., trees that were measured for the first time in the last census. These measures are underestimated for two reasons: (i) they do not account for branchfalls, which are an important part of biomass turnover in tropical forests and account for about a fourth of the area of canopy gaps in BCI [@Araujo2021], and (ii) some stems are recruited and die between two censuses, and are therefore never measured. A correction for the latter source of bias has been proposed by @Kohyama2019 and does not require additional measurements (applied in Figure \@ref(fig:fig2-correc)d).

* Complications of nonstandard height of measurement, data QAQC

The standard height for measuring tree diameter is 1.3 m (referred to as diameter at breast height, or DBH), and the vast majority of biomass allometries are calibrated with diameter values at 1.3 m. However, tree diameter is sometimes measured at a different height, for example because of trunk irregularity at 1.3 m height in buttressed trees; due to tree taper, this can lead to underestimation of tree diameter and thus biomass. Change in height of measurement, among other factors such as errors in DBH measurement and reporting, irregular or thorny trunks, and the presence of lianas or strangling figs can lead to outlier values for change in diameter [@Sheil1995]. The cause of these outliers is sometimes difficult to trace, making correction difficult, with multiple possible correction procedures all subject to bias [@Muller-Landau2014]. Nonstandard measurement height can be handled by multiplying the measured diameter by a "taper correction" factor; such a model has been calibrated with BCI data [@Cushman2014; see "taper correction" in Figure \@ref(fig:fig2-correc)]. To deal with outlier variation in diameter, outlier values can be replaced with the average growth of stems with similar characteristics (e.g., similar diameter; see "outlier substitution" in Figure \@ref(fig:fig2-correc)). 

Estimates of biomass stocks and fluxes at the plot level are very sensitive to the choice of biomass allometry (Figure \@ref(fig:fig1-allom)) and to the procedures used to deal with non-standard measurement heights and outliers for diameter changes (Figure \@ref(fig:fig2-correc)). In addition, there are specific problems with the early BCI plot censuses: in the first census, all trees were measured at breast height, even buttress trees with highly irregular trunks; therefore, we have removed the first census (1981-1982) from the figures in this chapter. In the first two censuses, the diameter of small stems (< 5.5 cm) was rounded to the nearest 5 mm, while the diameter of all other stems was reported to the nearest 1 mm; in the last censuses, all stem diameters were reported to 1 mm. Finally, prior to 2010, there were no individual stem tags on multi-stemmed trees, making it difficult to estimate their biomass dynamics. Together, these factors explain why no two papers give the same values for BCI biomass stocks and fluxes [e.g., @Chave2003;@Rutishauser2020 xxx].


```{r fig2-correc, fig.height = 8, fig.width = 12, fig.cap="Some examples of corrections on individual tree aboveground biomass (AGB; Mg) and AGB changes (Mg/yr), and their effect on total plot AGB (Mg/ha) and aboveground woody productivity (AWP; Mg/ha/yr)."}
# Figure 2 - Corrections ####
# fig2a - illustration with one problematic tree
df_ind <- subset(df_ind, stemID=="1428")

df_ind$agb <- df_ind$chave14
# could substitute dbh with agb
vars <- paste0(rep(c("D", ""), each = 3), "agb", c("", "_t", "_ts"))
df_ind_melt <- data.table::melt(df_ind, measure.vars = vars, id.vars = "year")

df_ind_melt[, method := data.table::tstrsplit(variable, "_")[[2]]]
df_ind_melt[is.na(method), method := "original"]
df_ind_melt[, variable := data.table::tstrsplit(variable, "_")[[1]]]

# change name of methods and variable levels
df_ind_melt$method <- factor(df_ind_melt$method, levels = c("original", "t", "ts"))
levels(df_ind_melt$method) <- list("Original data" = "original",
                                   "Taper correction" = "t", 
                                   "Taper correction + Outlier substitution" = "ts")

df_ind_melt$variable <- factor(df_ind_melt$variable, levels = c("agb", "Dagb"))
levels(df_ind_melt$variable) <- list("AGB (Mg)" = "agb", "AWP (Mg/yr)" = "Dagb")

# make panel labels 
dflabs <- data.frame(variable = levels(df_ind_melt$variable), label = letters[1:2])

# make individual level panels a and b
fig2ab_corr_tree <- ggplot(df_ind_melt, aes(x = year, y = value, color = method)) +
  geom_line(lty = 2) +
  geom_point() + 
  geom_line(data = subset(df_ind_melt, method=="Taper correction"), lty = 2) +
  geom_line(data = subset(df_ind_melt, method=="Original data"), lty = 2) +
  geom_point(data = subset(df_ind_melt, method=="Taper correction")) +
  geom_point(data = subset(df_ind_melt, method=="Original data")) +
  geom_hline(yintercept = 0, lty = 3, variable = "AWP (Mg/yr)") +
  geom_text(data = dflabs, aes(label = label), size = 6, hjust = 4, vjust = 1,
            x= -Inf, y = Inf, inherit.aes = FALSE) +
  coord_cartesian(clip = "off") +
  expand_limits(y = 0) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(y = "", x = "", title = paste0("Individual tree level (tree #", df_ind$treeID[1], " - ", df_ind$Latin[1], ")")) +
  scale_color_manual(values = c("black", "#1F78B4", "#E31A1C")) +
  theme_classic() +
  theme(strip.placement = "outside", 
        strip.background = element_blank(), 
        legend.position = "none",
        plot.title = element_text(hjust = 0.5, vjust = 2, size = 15)) +
  facet_wrap(~variable, scales = "free", strip.position = "left")

# fig2b - corrections on plot AGB
dfcorr <- subset(df_plot, !grepl("chave05|_h", method) & !(variable=="awp" & year == 2015))

# put awp points between the 2 years that it's based on
dfcorr[variable == "awp", year := year + 2.5]

# change order of method and variable levels
dfcorr$method <- factor(dfcorr$method, levels = unique(dfcorr$method))
levels(dfcorr$method) <- list("Original data" = "chave14", 
                              "Taper correction" = "chave14+taper", 
                              "Taper correction + Outlier substitution" = "chave14+taper+subs", 
                              "Original data+ Kohyama correction" = "chave14+kohyama",
                              "Taper correction+ Kohyama correction" = "chave14+taper+kohyama",
                              "Taper correction+ Outlier substitution\n+ Kohyama correction" = "chave14+taper+subs+kohyama")


dfcorr$variable <- factor(dfcorr$variable, levels = c("agb", "awp"))
levels(dfcorr$variable) <- list("AGB (Mg/ha)" = "agb", "AWP (Mg/ha/yr)" = "awp")

# make panel labels 
dflabs <- data.frame(variable = levels(dfcorr$variable), label = letters[3:4])

fig2cd_corr_plot <-
  ggplot(dfcorr, aes(x = year, y = value, color = method)) +
  geom_point() + geom_line(lty = 2) +
  geom_line(data = subset(dfcorr, method=="Original data"), lty = 2) +
  geom_point(data = subset(dfcorr, method=="Original data")) +
  theme_classic() +
  geom_text(data = dflabs, aes(label = label), size = 6, hjust = 4, vjust = 1,
            x= -Inf, y = Inf, inherit.aes = FALSE) +
  coord_cartesian(clip = "off") +
  labs(x = "Census year", y = "", colour = "Correction applied", title = "Plot level") + 
  theme(strip.placement = "outside", 
        strip.background = element_blank(),
        plot.title = element_text(hjust = 0.5, vjust = 2, size = 15)) +
  scale_color_manual(values = c("black", "#1F78B4", "#E31A1C", "grey", "#A6CEE3", "#FB9A99")) +
  facet_wrap(~variable, scales = "free", strip.position = "left") +
  expand_limits(y = 0) +
  scale_y_continuous(expand = c(0, 0))

fig2_leg <- ggpubr::as_ggplot(ggpubr::get_legend(fig2cd_corr_plot))

plts <- ggpubr::ggarrange(fig2ab_corr_tree, 
                          fig2cd_corr_plot + theme(legend.position = "none"), 
                          nrow = 2)
ggpubr::ggarrange(plts, fig2_leg, nrow = 1, widths = c(4, 1))
```


```{r fig3-groups, fig.height = 4, fig.width = 8, fig.cap="Stacked barplots showing the proportion (%) of aboveground biomass (AGB) and aboveground woody productivity (AWP) in different groups. (a) Proportion of AGB (left) and AWP (right) by size class (from blue to red).  Proportion of AGB (left) and AWP (right) by plant functional type as defined in @Ruger2020."}
# Figure 3 - AGB per group ####

# # estimate percentage of each size class or PFT group per variable
df_size[, `:=`(agb = agb/sum(agb), awp = awp/sum(awp))]
# revert order of size classes to have bigger trees above
df_size$size <- factor(df_size$size, levels = rev(levels(df_size$size)))
# 
df_pft[, `:=`(agb = agb/sum(agb), awp = awp/sum(awp))]

df_size_bis <- data.table::melt(df_size, measure.vars = c("agb", "awp"))
# change variable labels (AGB, AWP) to upper case
levels(df_size_bis$variable) <- toupper(levels(df_size_bis$variable))
# figure 3a bis
fig3a_bis_groups_size <- ggplot(df_size_bis) + 
  geom_col(aes(x = variable, y = value*100, fill = size)) + 
  labs(x = "", y = "Proportion (%)", fill = "Diameter\nclass (cm)") +
  scale_y_continuous(expand = c(0, 0)) +  
  scale_fill_brewer(palette = "RdYlBu") +
  theme_classic()

df_pft_bis <- data.table::melt(df_pft, measure.vars = c("agb", "awp"))
# change variable labels (AGB, AWP) to upper case
levels(df_pft_bis$variable) <- toupper(levels(df_pft_bis$variable))
# figure 3a bis
fig3b_bis_groups_pft <- ggplot(df_pft_bis) + 
  geom_col(aes(x = variable, y = value*100, fill = PFT)) + 
  labs(x = "", y = "Proportion (%)", fill = "PFT") +
  scale_y_continuous(expand = c(0, 0)) +  
  scale_fill_manual(values = c("gold", "darkorange", "seagreen", "royalblue", "orchid")) +
  theme_classic()

ggpubr::ggarrange(fig3a_bis_groups_size, fig3b_bis_groups_pft, ncol = 2, labels = "auto")
```


# Plot-based tree aboveground biomass results 

* Temporal variation in BCI woody tree AGB stocks and fluxes 

The fate of carbon storage in tropical forest biomass under climate change is still uncertain. Factors such as increased availability of atmospheric carbon dioxide may increase tree productivity, but there is also growing evidence that tree mortality is increasing, likely because of rising temperature and frequency of extreme droughts [@Hubau2020]. Permanent tropical forest plots such as BCI, which provide a great temporal depth of data on forest biomass stocks and fluxes (productivity, mortality), are therefore essential to assess whether these stocks and fluxes are increasing or decreasing over time, and to evaluate their relationship to climate. Estimates and temporal trends of BCI biomass stocks and fluxes varied over time depending on the allometric equations used and the data correction procedure (Table 1, Figure \@ref(fig:fig1-allom), Figure \@ref(fig:fig2-correc)). Overall, there is no evidence of a systematic directional change in BCI biomass stocks and fluxes, although biomass stocks and fluxes show strong temporal variation [Figure \@ref(fig:fig1-allom); @Feeley2007;@Rutishauser2020].

* Contributions to tree AGB stocks and fluxes by subgroups

Tree characteristics (size, canopy stature, growth rate, etc.) provide insight into their ecological role in a forest. In particular, these characteristics can influence their relative contribution to biomass stocks and fluxes. For example, it has been shown that small trees contribute relatively more to productivity than to total biomass [Figure \@ref(fig:fig3-groups); @Piponiot2021;@Meakem2018]. xx a few things about the functional groups based on the one we chose for Figure 3.

*	Spatial variation in BCI AGB stocks and fluxes – within BCI 50 ha plot 

Biomass stocks and fluxes exhibit high small-scale spatial variability in the 50 ha BCI plot [@Rejou-Mechain2014]. This high spatial noise in the data makes it difficult to identify clear patterns in the spatial distribution of biomass within the 50 ha plot (Figure \@ref(fig:fig4-spatial)). Topography explains little variation in aboveground biomass stocks across the plot [@Ledo2016], but some differences across geological substrates have been detected [@Chave2003;@DeWalt2004; Supplementary material?]. Liana abundance has been shown to be negatively correlated with aboveground biomass [@Ledo2016; ref chapters on liana removal experiments, on ecology of lianas]. Finally, a small portion of the 50-ha plot is composed of secondary forest (outline in future Figure 4) and has thus lower aboveground biomass. 

*	Spatial variation in BCNM AGB stocks and fluxes from multiple plots, and experimental results 

The 50 ha plot is not representative of the island of Barro Colorado, let alone the entire Barro Colorado National Monument: it is located in a particularly flat part of the island, with a relatively homogeneous geological substrate and fairly old forest [@Baillie2006;@Mascaro2011]. This contrasts with the rest of the island, which has a greater diversity of soil, topography, and forest age [@Baillie2006;@Mascaro2011]. To understand how biomass stocks and fluxes vary across Barro Colorado National Monument, smaller forest plots were set up: xx large plots spread across the island where only the largest trees (>20 cm DBH) were measured, and xxx 1 ha plots where all trees larger than 10 cm DBH were measured. (map in supplementary material? + figure or table showing AGB stocks and fluxes variation?)

    + The 50 ha plot has a larger mean biomass then the rest of the Barro Colorado National Monument (ref figure with 1-ha plots + large BCI plots). 

    + Plot data also used to calibrate remote sensing and these then used to estimate broader spatial patterns – ref other chapter on this.

*	Experimental results from BCNM work on AGB stocks and fluxes (one sentence on each, and refs to associated chapters)

    + Soil fertility – Gigante fertilization experiments 

    + Litter addition/removal

    + Liana removal 

> Other woody biomass pools and fluxes ?

```{r, fig4-spatial, fig.height = 8, fig.width = 12, fig.cap="Crown-distributed AGB (Mg/ha) and AWP (Mg/ha/yr) on the 50 ha plot, grouped by topographically-defined habitat [as defined by @Harms2001]. (a) Map of crown-distributed AGB in the 2010 census. (b)  Map of crown-distributed AWP between the 2010 and 2015 censuses. Colors delineate the topographically defined habitats. (c) Mean (+/- standard error across 20 m quadrats) of AGB and AWP in each topographically defined habitat. The AGB and AWP of individual trees were distributed among 2x2m pixels in proportion to the area of intersection between the pixel and the tree crown [crown radius calculated according to @MartinezCano2019]. The AGB and AWP of the portions of tree crowns that lie outside the edges were mirror transposed into the pixels inside the plot to avoid underestimation of AGB and AWP at the edges of the plot (see Supplementary material). Results are consistent with a similar analysis by @Chave2003."}

load("data/data-main-script.rda")
df_crown$agb[df_crown$agb>1000] <- 1000
df_crown$awp[df_crown$awp>30] <- 30

## get bci habitats from Harms et al 2001
load("data/bci_habitat.rda")
# change name of habitat to match publication: first letter in upper case and add spaces
levels(bci_habitat$habitat) <- paste0(toupper(substr(levels(bci_habitat$habitat), 1, 1)), 
                              substr(levels(bci_habitat$habitat), 2, nchar(levels(bci_habitat$habitat))))
levels(bci_habitat$habitat) <- gsub("Hi_", "High ", levels(bci_habitat$habitat))
levels(bci_habitat$habitat) <- gsub("_", " ", levels(bci_habitat$habitat))
levels(bci_habitat$habitat) <- gsub("Young", "Young forest", levels(bci_habitat$habitat))

## Convert bci_habitat to a raster object
bci_habitat[, habitat_f := as.numeric(as.factor(habitat))]
bci_habitat$secondary = bci_habitat$habitat=="young"
## translate x and y to have center of quadrats
bci_habitat$x = bci_habitat$x+10; bci_habitat$y = bci_habitat$y + 10
r <- raster::rasterFromXYZ(bci_habitat[, c("x", "y", "habitat_f")])

## Extract polygons and make it a simple feature 
habitats <- raster::rasterToPolygons(r, dissolve = TRUE) 
habitats <- sf::st_as_sf(habitats) 
habitats <- merge(habitats, unique(bci_habitat[, c("habitat", "habitat_f")]))
habitats_buff <- sf::st_buffer( habitats, -2)
           
gagb <- ggplot(df_crown) + 
  geom_raster(aes(x = x, y = y, fill = agb)) +
  scale_fill_gradient(low = "black", high = "white", 
                      breaks = c(0, 250, 500, 750, 1000),
                      labels = c(0, 250, 500, 750, ">1000")) +
  labs(fill = "AGB\n(Mg/ha)", x = "", y = "") +
  expand_limits(x = c(0, 1000), y = c(0, 500)) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_bw()
# isolate legend
legend1 <- ggpubr::as_ggplot(ggpubr::get_legend(gagb))
# remove legend from gagb and add the habitat layer
gagb <- gagb + 
  geom_sf(data = habitats_buff, aes(color = habitat), size = 1.4, fill = NA) +
  theme(legend.position = "none")
  
gcolor <- ggplot(habitats_buff, aes(color = habitat)) + geom_sf(fill = NA, size = 1.5)
legend2 <- ggpubr::as_ggplot(ggpubr::get_legend(gcolor))
  
gawp <- ggplot(df_crown) + 
  geom_raster(aes(x = x, y = y, fill = awp))  +
  scale_fill_gradient(low = "black", high = "white",
                      breaks = c(0, 10, 20, 30),
                      labels = c(0, 10, 20, ">30")) +
  labs(fill = "AWP\n(Mg/ha/yr)", x = "", y = "") +
  expand_limits(x = c(0, 1000), y = c(0, 500)) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_bw()
# isolate gawp legend
legend3 <- ggpubr::as_ggplot(ggpubr::get_legend(gawp))
gawp <- gawp + theme(legend.position = "none")+
  geom_sf(data = habitats_buff, aes(color = habitat), size = 1.4, fill = NA)

## stats by habitat
# add habitat information to df_crown
# 1. add quadrat based on x and y
df_crown[, `:=` (q1 = as.character(floor(x/20)), q2 = as.character(floor(y/20)))]
df_crown[nchar(q2) == 1, q2 := paste0("0", q2)]
df_crown[, q20 := as.numeric(paste0(q1, q2))]
df_crown[, `:=`(q1 = NULL, q2 = NULL)]
# merge with bci_habitats
df_crown <- merge(df_crown, bci_habitat[, c("q20", "habitat")], by = "q20")

dfhab2 <- data.table::melt(df_crown, measure.vars = c("agb", "awp"), id.vars = c("habitat", "q20"))
# group by quadrat
dfhab2 <- dfhab2[, .(value = mean(value)), .(variable, q20, habitat)]

dfhab2 <- dfhab2[, .(mean = mean(value), se = sd(value)/sqrt(length(value))), .(habitat, variable)]
# levels(dfhab2$variable) <- c("AGB (Mg/ha)", "AWP (Mg/ha/yr)")

gagb2 <- ggplot(subset(dfhab2, variable == "agb"), 
                aes(x=habitat, y = mean, ymin = mean-se, ymax = mean+se, col = habitat)) +
  geom_pointrange() +
  labs(x = "", y = "AGB (Mg/ha)") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1), 
        legend.position = "none")

gawp2 <- ggplot(subset(dfhab2, variable == "awp"), 
                aes(x=habitat, y = mean, ymin = mean-se, ymax = mean+se, col = habitat)) +
  geom_pointrange() +
  labs(x = "", y = "AWP (Mg/ha/yr)") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1), 
        legend.position = "none")

# group graphs
gmap <- ggpubr::ggarrange(gagb, gawp, nrow=2, labels = c("a","b"))
gbar <- ggpubr::ggarrange(gagb2, gawp2, nrow=2, labels = c("c", "d"))
legends <- ggpubr::ggarrange(legend1, legend2, legend3, nrow=3)

ggpubr::ggarrange(gmap, legends, gbar, widths = c(4,1,2), ncol=3)
```

# Supplementary material




# References
